Exploring the Relationship between Family Background and Children's Education Level in the United States (2021)

### 2023-04-17 outputs/paper/paper.qmd
final draft of the paper

In the past, family status variables, such as the socio-economic status and educational level of parents, were commonly considered as strong predictors of academic achievement in children. More specifically, parents who have lower levels of education, and those with higher levels of education are more inclined to view higher education as desirable and encourage their children to excel academically. They also tend to hold higher expectations for their children's academic performance [@davis_2014]. Ultimately, these may affect the education level of the children.

One potential explanation for this relationship is that higher levels of education can provide parents with access to resources such as income, time, energy, and community contacts, which can enable greater involvement in a child's education. As a result, the influence of family status variables on children's academic achievement may be best understood as a complex interaction between status and process variables [@khan].

However, recent research has indicated that the relationship between these factors and academic achievement is not always direct. Instead, socio-economic status and parents' education are part of a larger set of psychological and sociological variables that can impact children's educational outcomes [@khan].

@@

@tb-regandeduc showcases the average number of years of education for individuals in different living regions of the United States when the respondent was 16. The highest average education level is found in the Middle Atlantic region, where the mean is 15.13 years. The East North Central region follows closely, with an average of 14.98 years of education. In the Pacific and New England regions, the mean years of education are relatively similar, with 14.94 and 14.85 years, respectively. Those regions are above the overall average, which is 14.82 from @tb_educ. On the other hand, the South Atlantic and West North Central regions also display comparable average education levels of 14.75 and 14.73 years, respectively. The Mountain region exhibits a slightly lower mean of 14.69 years. Furthermore, the East South Atlantic and West South Central regions have the lowest average education levels, with 14.10 and 14.39 years, respectively.

@@

modelsummary::modelsummary(list("Children's Education" = model),
                           fmt = 2,
                           statistic = "conf.int")
```

The @tbl-lmeduc is built based on the @eq-model. In this logistic model, it is used to predict the probability of a child attains a post-secondary degree based on these predictors.
*missing the bullet point to explain each exp(coef)*

From @tbl-lmeduc, the positive coefficients for both mother's and father's prestige scores (**mapres10** and **papres10**) suggest a slight positive relationship with children's education levels. Moreover, the positive coefficients for parents' education (**paeduc** and **maeduc**) have a greater positive association with children's education level.

The coefficients for the income level at age of 16 (**incom16**) indicate that higher-income levels are generally associated with lower education levels for children. Particularly, the "far above average" and "far below average" income categories show a significant negative relationship with children's education levels.

Regarding the living regional differences at age of 16 (**reg16**), the results indicate that children from the east south atlantic region may have lower education levels compared to those from other regions. The positive coefficients for middle atlantic and west north central regions suggest a positive association with children's education levels, although the results are not statistically significant.

@@

@@

The correlation between parents' education levels and their children's educational outcomes is well-established in the literature. Studies have consistently shown that parents' education is one of the strongest predictors of their children's academic achievement and educational attainment [@hout]. @chetty used data from tax records and college enrollment records to analyse the relationship between parents' income, education, and social status, and their children's earnings and educational outcomes. The results showed that children from high-income families were more likely to attend college and earn higher incomes than children from low-income families. In addition, the study found that parental education was a key factor in determining children's educational and economic outcomes. Children with highly educated parents were more likely to attend college and earn higher incomes than children with less educated parents.

Furthermore, the analysis of the statistical model from the @tbl-lmeduc shows that parental education and prestige scores are positively correlated with their children's academic achievement and societal status. The coefficients for the independent variables father's and mother's education in years indicate that an increase in fathers' education has a greater association with their children's education than that of mothers. This suggests that the father's role in education and investment in their children's future may be particularly important. However, the influence of both parents is still significant. The R-squared value suggests that the independent variables explain only a small portion of the variation in the dependent variable, indicating that other factors beyond parental education and prestige score also play a role. Nonetheless, these findings highlight the importance of parental involvement in their children's education, with a particular emphasis on the role of fathers.

The relationship between parental education and student achievement has been the focus of many studies, with varying results. Some study has found that maternal educational attainment had an impact on student achievement, which contradicts the finding in this paper [@1993]. The present study showed that paternal support of education has a significant impact on academic achievement, while maternal education level did not show a significant impact. However, the study of @2000 showed that families with active fathers fostered maternal involvement in the household, which resulted in a support system that fosters educational attainment and positive long-term contributions.

Despite these discrepancies, the findings in this paper concur with prior literature that emphasizes the importance of parental education levels and their impact on college student achievement [@ortiz]. This study focuses on showing that the fathers' education level has a significant relationship with childrens' year of schooling, while no statistically significant relationship was observed with the mothers' education level.

@@

## The Effect of Living Region during the Adolescent Period
referring results from the model about how regions affect the education level

>>>>

Education plays a crucial role in shaping an individual's life chances, personal development, and socioeconomic mobility. Over the years, researchers have sought to understand the factors that may influence children's academic achievement and education levels, with a particular focus on the impact of family background variables such as parents' socio-economic status and educational attainment [@coleman; @sirin]. To reveal the complex relationship between family background and children's education, it is important to developing effective educational policies and interventions that promote equal opportunities for all students, regardless of their social or economic origins.

Several studies have documented the strong and consistent association between parents' education and children's academic outcomes [@haveman]. Parents with higher education levels are more likely to provide an intellectually stimulating home environment, hold higher expectations for their children's academic performance, and be actively involved in their education, thus fostering their children's academic success [@davis2]. However, recent research has highlighted the multifaceted nature of this relationship, indicating that it can be influenced by various factors such as family income, social class, and so on [@reardon, @domina].

Despite the substantial body of research on this topic, there is still much to learn about the intricate ways in which family background affects children's education levels, particularly in the context of growing economic inequality and social stratification [@chetty2]. This paper aims to contribute to the ongoing scholarly discourse by examining the relationship between family status variables, including parents' socio-economic status and educational attainment, and their impact on children's academic achievement and education level using data from the 2021 General Social Survey (GSS). By gaining a deeper understanding of these dynamics,policymakers will be better equipped to address educational disparities and work toward fostering more equitable opportunities for all students.

@@

@tbl-regandeduc showcases the average number of years of education for individuals in different living regions of the United States when the respondent was 16. The highest average education level is found in the Middle Atlantic region, where the mean is 15.13 years. The East North Central region follows closely, with an average of 14.98 years of education. In the Pacific and New England regions, the mean years of education are relatively similar, with 14.94 and 14.85 years, respectively. Those regions are above the overall average, which is 14.82 from @tbl-educ. On the other hand, the South Atlantic and West North Central regions also display comparable average education levels of 14.75 and 14.73 years, respectively. The Mountain region exhibits a slightly lower mean of 14.69 years. Furthermore, the East South Atlantic and West South Central regions have the lowest average education levels, with 14.10 and 14.39 years, respectively.

@@

modelsummary::modelsummary(list("Children's Education" = model),
                           fmt = 2,
                           statistic = "conf.int")
```
$$
\mbox{Odds}=\frac{\pi}{1-\pi}=\mbox{exp}(\hat{\beta_0}+\hat{\beta_1}X_{\mbox{mapres10}}+\hat{\beta_2}X_{\mbox{papres10}} + \hat{\beta_3}X_{\mbox{paeduc}} + \hat{\beta_4}X_{\mbox{maeduc}})
$$ {#eq-coef}

The @tbl-lmeduc is built based on the @eq-model. In this logistic model, it is used to predict the probability of a child attains a post-secondary degree based on these predictors.

Using the @eq-coef, the coefficients of the model can be interpreted as follow and note that the base group for income level is `Average` and for region is `East North Central`:

- **Intercept**: The odds of a child completing high school when all other predictor variables are at their reference levels (i.e., income level is 'average' and region is 'east north') is exp(-2.59) = 0.075, with a 95% confidence interval of [exp(-2.99), exp(-2.19)] = [0.05, 0.11].

- **Respondent's Mother's Prestige Score (mapres10)**: A one-unit increase in the respondent's mother's prestige score is associated with an increase in the odds of a child completing high school by exp(0.01) = 1.01, with a 95% confidence interval of [exp(0.01), exp(0.02)] = [1.01, 1.02].

- **Respondent's Father's Prestige Score (papres10)**: A one-unit increase in the respondent's father's prestige score is associated with an increase in the odds of a child completing high school by exp(0.02) = 1.02, with a 95% confidence interval of [exp(0.01), exp(0.02)] = [1.01, 1.02].

- **Respondent's Father's Education (paeduc)**: A one-year increase in the number of years of education completed by the respondent's father is associated with an increase in the odds of a child completing high school by exp(0.10) = 1.11, with a 95% confidence interval of [exp(0.07), exp(0.12)] = [1.07, 1.13].

- **Respondent's Mother's Education (maeduc)**: A one-year increase in the number of years of education completed by the respondent's mother is associated with an increase in the odds of a child completing high school by exp(0.03) = 1.03, with a 95% confidence interval of [exp(0.01), exp(0.06)] = [1.01, 1.06].

- **Respondent's Income Level above Average (incom16above average)**: The odds of a child completing high school for respondents with an above average household income at age 16 are exp(0.29) = 1.34 times the odds for respondents with an average household income, with a 95% confidence interval of [exp(0.09), exp(0.50)] = [1.09, 1.65].

- **Respondent's Income Level below Average (incom16below average)**: The odds of a child completing high school for respondents with a below-average household income at age 16 are exp(-0.11) = 0.90 times the odds for respondents with an average household income, with a 95% confidence interval of [exp(-0.28), exp(0.06)] = [0.75, 1.06].

- **Respondent's Income Level Far above Average (incom16far above average)**: The odds of a child completing high school for respondents with a far above-average household income at age 16 are exp(0.28) = 1.33 times the odds for respondents with an average household income, with a 95% confidence interval of [exp(-0.20), exp(0.78)] = [0.82, 2.18].

- **Respondent's Income Level Far below Average (incom16far below average)**: The odds of a child completing high school for respondents with a far below-average household income at age 16 are exp(-0.21) = 0.81 times the odds for respondents with an average household income, with a 95% confidence interval of [exp(-0.44), exp(0.03)] = [0.64, 1.03].

- **Respondent's Living Region in East South Atlantic at 16 (reg16east south atlantic)**: The odds of a child completing high school for respondents living in the East South Atlantic region at age 16 are exp(-0.18) = 0.83 times the odds for respondents living in the East North region, with a 95% confidence interval of [exp(-0.49), exp(0.14)] = [0.61, 1.15].

- **Respondent's Living Region in Middle Atlantic at 16 (reg16middle atlantic)**: The odds of a child completing high school for respondents living in the Middle Atlantic region at age 16 are exp(0.26) = 1.30 times the odds for respondents living in the East North region, with a 95% confidence interval of [exp(0.02), exp(0.50)] = [1.02, 1.65].

- **Respondent's Living Region in Mountain at 16 (reg16mountain)**: The odds of a child completing high school for respondents living in the Mountain region at age 16 are exp(-0.19) = 0.83 times the odds for respondents living in the East North region, with a 95% confidence interval of [exp(-0.51), exp(0.12)] = [0.60, 1.13].

- **Respondent's Living Region in New England at 16 (reg16new england)**: The odds of a child completing high school for respondents living in the New England region at age 16 are exp(0.09) = 1.09 times the odds for respondents living in the East North region, with a 95% confidence interval of [exp(-0.26), exp(0.45)] = [0.77, 1.57].

- **Respondent's Living Region in Pacific at 16 (reg16pacific)**: The odds of a child completing high school for respondents living in the Pacific region at age 16 are exp(0.02) = 1.02 times the odds for respondents living in the East North region, with a 95% confidence interval of [exp(-0.20), exp(0.24)] = [0.82, 1.27].

- **Respondent's Living Region in South Atlantic at 16 (reg16south atlantic)**: The odds of a child completing high school for respondents living in the South Atlantic region at age 16 are exp(-0.06) = 0.94 times the odds for respondents living in the East North region, with a 95% confidence interval of [exp(-0.30), exp(0.18)] = [0.74, 1.20].

- **Respondent's Living Region in West North Central at 16 (reg16west north central)**: The odds of a child completing high school for respondents living in the West North Central region at age 16 are exp(-0.16) = 0.85 times the odds for respondents living in the East North region, with a 95% confidence interval of [exp(-0.43), exp(0.12)] = [0.65, 1.13].

- **Respondent's Living Region in West South Central at 16(reg16west south central)**: The odds of a child completing high school for respondents living in the West South Central region at age 16 are exp(-0.33) = 0

@@

## The impact of family income level and prestige score

From @tbl-lmeduc, it show the coefficients for the income level at age of 16 and indicates that higher-income levels are generally associated with lower education levels for children. Particularly, the "far below average" income categories show a significant negative relationship with children's education levels in terms of the odds of attaining a post-secondary degree.

The correlation between parents' education levels and their children's educational outcomes is well-established in the literature. Studies have consistently shown that parents' education is one of the strongest predictors of their children's academic achievement and educational attainment [@hout]. @chetty used data from tax records and college enrollment records to analyse the relationship between parents' income, education, and social status, and their children's earnings and educational outcomes. The results showed that children from high income families were more likely to attend college and earn higher incomes than children from low income families. The findings from the paper match with the existing research, that higher income level has a postive effect on the children's education level.

However, the coefficients for parents' (both mother's and father's) prestige score show a neglectable positive effect. The effect is too small to draw an remarkable conclusion. Yet, the purpose of using the prestige score is to measure the family socialeconomic status based on their occupation. There are some reason to explain why the effects are not significant and @sec-weakness will provide some potential solution to better examine this relationship.

@@

Furthermore, the analysis of the logistic model from the @tbl-lmeduc shows that parental education and prestige scores are positively correlated with their children's probability of attaining a post-secondary degree. The coefficients for the independent variables father's and mother's education in years indicate that an increase in fathers' education has a greater association with their children's education than that of mothers. This suggests that the father's role in education and investment in their children's future may be particularly important. However, the influence of both parents is still significant. Nonetheless, these findings highlight the importance of parental education level in their children's education, with a particular emphasis on the role of fathers.

The relationship between parental education and student achievement has been the focus of many studies, with varying results. Some study has found that maternal educational attainment had an impact on student achievement, which contradicts the finding in this paper [@1993]. The present study showed that paternal support of education has a significant impact on academic achievement, while maternal education level did not show a significant impact. Additionally, the study of @2000 showed that families with active fathers fostered maternal involvement in the household, which resulted in a support system that fosters educational attainment and positive long-term contributions. This suggest that the role of parents both contribute to children's education.

The findings in this paper concur with prior literature that emphasizes the importance of parental education levels and their impact on college student achievement [@ortiz]. This study focuses on showing that the fathers' education level has a significant relationship with childrens' year of schooling, while no  significant relationship was observed with the mothers' education level.

In addition, the study found that parental education was a key factor in determining children's educational and economic outcomes. Children with highly educated parents were more likely to attend college and earn higher incomes than children with less educated parents [@chetty].

## The Effect of Living Region during the Adolescent Period

Regarding the living regional differences at age of 16 from @tbl-lmeduc, the coefficients indicate that children from the East South Atlantic, West South Central, Mountain and West North Central may have a negative effect on completed a post-secondary degree compared to those from other regions. The positive coefficients for Middle Atlantic regions and New England suggest a positive association with children's education levels, although the results are not statistically significant. The @eq-model employs the probability of a child obtaining a post-secondary degree as the response variable. A potential explanation for the positive coefficient could be that certain regions offer higher-quality education, particularly in the form of high schools.

According to the US News rankings in 2022, these are the top 8 high schools in the US. Thomas Jefferson High School for Science and Technology is located in the South Atlantic region in Virginia, as is the Academic Magnet High School in South Carolina and the School for Advanced Studies (SAS) in Florida. The Signature School can be found in the East North Central region in Indiana, while the Payton College Preparatory High School is also located in the East North Central region in Illinois. The Davidson Academy of Nevada is situated in the Mountain region in Nevada. The Central Magnet School is in the East South Central region in Tennessee, and The School for the Talented and Gifted (TAG) is located in the West South Central region in Texas. [@best]

It's noteworthy that none of the top schools are from the Middle Atlantic or New England regions, which have positive coefficients in the model. This discrepancy could be due to differences in time. School rankings can change over the years as states develop, and the GSS respondents were born in a different time period. This leads to a mismatch between the rankings and the model's coefficients.

@@
